Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-22 main.tex
syntax
some more stuff
more semantics
Update on Overleaf.
fix
more rules ugh

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
%\usepackage{amssymb}
\usepackage{array}
\usepackage{minted}
\usepackage{ifthen}

% Configuration for minted package
\setminted{
    style=colorful, % Set code style to colorful
    fontsize=\scriptsize,
    linenos, % Enable line numbers
    breaklines, % Enable line breaks
    breakanywhere, % Allow line breaks anywhere
    autogobble,
}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}

\ifthenelse{1=1}{
\newcommand{\TODO}[1]{\textcolor{red}{TODO: #1}}
\newcommand\help[1]{\textcolor{green}{Help Needed: #1}}
\newcommand\teo[1]{\textcolor{red}{teo: #1}}
\newcommand\willow[1]{\textcolor{purple}{Willow: #1}}
\newcommand{\saman}[1]{\textcolor{orange}{Saman: #1}}
\newcommand{\leftoff}{\textcolor{red}{\textbf{LEFT OFF}}}
\newcommand{\changwan[1]{\textcolor{blue}{CW: #1}}
}{
\newcommand{\TODO}[1]{}
\newcommand{\help}[1]{}
\newcommand{\saman}[1]{}
\newcommand{\teo}[1]{}
\newcommand{\willow}[1]{}
\newcommand{\leftoff}{}
}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Finch: A Datastructure-Driven Array Programming Language}

@@

Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes (from Fred's thesis!). So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
 Thus, focusing only on sparsity limits the scope needed to handle a vast array of data. Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.

>>>>

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
%\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{array}
\usepackage{minted}
\usepackage{ifthen}
\usepackage{ebproof}
\ebproofset{label separation=0pt}
\usepackage{multirow}

% Configuration for minted package
\setminted{
    style=colorful, % Set code style to colorful
    fontsize=\scriptsize,
    linenos, % Enable line numbers
    breaklines, % Enable line breaks
    breakanywhere, % Allow line breaks anywhere
    autogobble,
}
\setmintedinline{fontsize=\small}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}

\ifthenelse{1=1}{
\newcommand{\TODO}[1]{\textcolor{red}{TODO: #1}}
\newcommand\help[1]{\textcolor{green}{Help Needed: #1}}
\newcommand\kyle[1]{\textcolor{olive}{Kyle: #1}}
\newcommand\teo[1]{\textcolor{red}{teo: #1}}
\newcommand\willow[1]{\textcolor{purple}{Willow: #1}}
\newcommand{\saman}[1]{\textcolor{orange}{Saman: #1}}
\newcommand{\leftoff}{\textcolor{red}{\textbf{LEFT OFF}}}
\newcommand\changwan[1]{\textcolor{blue}{CW: #1}}
}{
\newcommand{\TODO}[1]{}
\newcommand{\help}[1]{}
\newcommand{\saman}[1]{}
\newcommand{\kyle}[1]{}
\newcommand{\teo}[1]{}
\newcommand{\willow}[1]{}
\newcommand{\leftoff}{}
}

\newcommand{\finchextent}{\ensuremath{\text{\textbf{extent}}}}
\newcommand{\finchliteral}{\ensuremath{\text{\textbf{literal}}}}
\newcommand{\finchvalue}{\ensuremath{\text{\textbf{value}}}}
\newcommand{\finchtensor}{\ensuremath{\text{\textbf{tensor}}}}
\newcommand{\finchmode}{\ensuremath{\text{\textbf{mode}}}}
\newcommand{\finchindex}{\ensuremath{\text{\textbf{index}}}}
\newcommand{\finchvar}{\ensuremath{\text{\textbf{variable}}}}
\newcommand{\finchcall}{\ensuremath{\text{\textbf{call}}}}
\newcommand{\finchaccess}{\ensuremath{\text{\textbf{access}}}}
\newcommand{\finchread}{\ensuremath{\text{\textbf{read}}}}
\newcommand{\finchassign}{\ensuremath{\text{\textbf{assign}}}}
\newcommand{\finchupdate}{\ensuremath{\text{\textbf{update}}}}
\newcommand{\finchloop}{\ensuremath{\text{\textbf{loop}}}}
\newcommand{\finchdefine}{\ensuremath{\text{\textbf{define}}}}
\newcommand{\finchsieve}{\ensuremath{\text{\textbf{sieve}}}}
\newcommand{\finchblock}{\ensuremath{\text{\textbf{block}}}}
\newcommand{\finchdeclare}{\ensuremath{\text{\textbf{declare}}}}
\newcommand{\finchfreeze}{\ensuremath{\text{\textbf{freeze}}}}
\newcommand{\finchthaw}{\ensuremath{\text{\textbf{thaw}}}}

\newcommand{\finchthunk}{\ensuremath{\text{\textbf{thunk}}}}
\newcommand{\finchphase}{\ensuremath{\text{\textbf{phase}}}}
\newcommand{\finchswitch}{\ensuremath{\text{\textbf{switch}}}}
\newcommand{\finchlookup}{\ensuremath{\text{\textbf{lookup}}}}
\newcommand{\finchrun}{\ensuremath{\text{\textbf{run}}}}
\newcommand{\finchspike}{\ensuremath{\text{\textbf{spike}}}}
\newcommand{\finchsequence}{\ensuremath{\text{\textbf{sequence}}}}
\newcommand{\finchstepper}{\ensuremath{\text{\textbf{stepper}}}}

\newcommand{\finchplan}{\ensuremath{\text{\textbf{plan}}}}
\newcommand{\finchquery}{\ensuremath{\text{\textbf{query}}}}
\newcommand{\finchtable}{\ensuremath{\text{\textbf{table}}}}
\newcommand{\finchmapjoin}{\ensuremath{\text{\textbf{mapjoin}}}}
\newcommand{\finchaggregate}{\ensuremath{\text{\textbf{aggregate}}}}
\newcommand{\finchrelabel}{\ensuremath{\text{\textbf{relabel}}}}
\newcommand{\finchreorder}{\ensuremath{\text{\textbf{reorder}}}}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Finch: Sparse and Structured Array Programming with Control Flow}

@@

Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes (from Fred's thesis!). So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
 Thus, focusing only on sparsity limits the scope needed to handle a vast array of data \kyle{"vast array" is a good pun here, but a bit confusing to read}. Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.

### 2024-03-22 sections/eval.tex
Update on Overleaf.

\subsection{Implementing Numpy Array API in Finch}

@@

@@

\subsubsection{Lowering}
\subsubsection{Heuristic Optimization}

>>>>

\subsection{Implementing Numpy Array API in Finch}
In the past decade, the adoption of the python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency. They generally either limit the dimensionality to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats. Further, existing work doesn't support the fusion of arbitrary operators which can have a drastic impact on performance as we show in Fig. (\kyle{PUT FIG HERE}). We believe that a flexible, online compiler like Finch which can manage the complexity of arbitrary operations between inputs with a wide variety of formats is the secret ingredient needed to make the array API performant for structured data.

However, there is a gap between the expressiveness of the purely declarative Array API and the Finch language which defines both outputs and the algorithm used to produce those outputs. The naive approach to solving this would be to explicitly define a mapping from each function in the API to a Finch program which eagerly produces the output. Unfortu

Due to the wide variety of formats and protocols supported by Finch, it allows for a large array of potential impleme

@@

\begin{align*}
    \finchreorder(name, idxs...) := 
\end{align*}

@@

\subsubsection{Lowering}

\subsubsection{Heuristic Optimization}
